---
title: "Data Manipulation"
output: github_document
date: "2026-09-24"
editor_options: 
  chunk_output_type: console
---

This file is for doing data manipulation

```{r}
library(tidyverse)
```

load litters and pups datasets

```{r}
litters_df =
  read_csv(
    "data/FAS_litters.csv",
    na = c("","NA","."))

litters_df = janitor::clean_names(litters_df)

pups_df =
  read.csv(
    "data/FAS_pups.csv",
    skip = 3,
    na = c("","NA","."))

pups_df = janitor::clean_names(pups_df)
```

select

```{r}
select(litters_df, group, litter_number, gd0_weight, pups_born_alive)

select(litters_df, group:gd_of_birth)

select(litters_df, -pups_survive)

select(litters_df, GROUP = group, LiTtEr_NuMbEr = litter_number)

rename(litters_df, GROUP = group, LiTtEr_NuMbEr = litter_number)

select(litters_df, starts_with("gd"))

select(litters_df, litter_number, pups_survive, everything())
```

alternatives

```{r}
rename(litters_df, GROUP = group)

relocate(litters_df, litter_number, gd0_weight)
```

in pups data, select litter number, sex, PD ears

```{r}
select(pups_df, litter_number, sex, pd_ears)
```

## `filter` some rows

```{r}
filter(litters_df, gd_of_birth == 19)

filter(litters_df, pups_born_alive <= 5)

filter(litters_df, pups_born_alive != 6)

filter(litters_df, group %in% c("Con7", "Con8"))

filter(litters_df, group %in% c("Con7", "Con8"), pups_born_alive < 7)

filter(litters_df, pups_born_alive < 4 | pups_born_alive > 8)

drop_na(litters_df, gd0_weight)
```

In the pups dataset, keep only rows where PD walk is less than 11 and sex is 2

```{r}
filter(pups_df, pd_walk < 11, sex == 2)
```

## `mutate` to change some columns

let's add and change some columns

```{r}
mutate(litters_df, 
       wt_gain = gd18_weight - gd0_weight,
       wt_gain_squared = wt_gain^2,
       group = str_to_lower(group))
```

in the pups dataset, create variable that subtracts 7 from PD walk

```{r}
mutate(pups_df, pd_walk_minus_7 = pd_walk - 7)
```

## `arrange`` data

```{r}
arrange(litters_df, pups_born_alive)
arrange(litters_df, desc(pups_born_alive))
```

## do multiple steps

```{r}
litters_df = 
  read_csv("data/FAS_litters.csv", na = c("","NA","."))|>
  janitor::clean_names() |>
  select(group, starts_with("gd")) |>
  drop_na() |>
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)
  )
```

load pups, clean names, drop missing, keep litter number and pd variables, add pd walk - 7

```{r}
pups_df =
  read_csv("data/FAS_pups.csv", skip = 3, na = c("","NA", ".")) |>
  janitor::clean_names() |>
  drop_na() |>
  select(litter_number, starts_with("pd")) |>
  mutate(
    pd_walk_minus_7 = pd_walk - 7
  )
```

sometimes the first thing isn't a dataframe

```{r}
litters_df |>
  filter(group %in% c("Con7", "Con8")) |>
  lm(gd18_weight ~ gd0_weight, data = _)
```

select vs pull

```{r}
litters_df |>
  select(group)

litters_df |>
  pull(group)
```